---
title: "Individual Project Reflection Report"
author: "Your name"
date: "Due Monday 11/25/2024"
format:
  html:
    toc: true
    toc_depth: 2
    number-sections: false
    number-depth: 2
    theme: yeti
    embed-resources: true
execute: 
  warning: false
  message: false
  echo: true
editor: 
  markdown: 
    wrap: 72
---



```{r,include=FALSE}
# These commands should always be near the top to ensure the libraries are loaded before the analysis.
library(mosaic)
```


# Introduction

This section provides a brief overview of the project, including its purpose, research questions, and key findings. Discuss why this project is relevant and what you aim to reflect on in this report.

# Personal Contribution

## Roles and Responsibilities
Describe your specific role in the project, detailing tasks you were responsible for and how you contributed to the overall group effort.

## Collaborative Efforts
Reflect on your teamwork experience, discussing both the challenges and successes of collaborating with your group.

# Key Statistical Findings

## Summary of Findings
Summarize 1-2 key findings from the project. Include an appropriate visualization or table below.

```{r}
# Insert your code here
# If you are accessing your data be sure you have included the input code above

```

## Methods Applied
Describe the statistical methods used in your analysis and discuss why they were suitable for your research questions.

# Personal Learning and Reflection

## Skills Acquired
Discuss the new skills and concepts you learned during the project, focusing on statistical techniques and R programming.

## Challenges Faced
Reflect on the difficulties you encountered and how you addressed them.

## Lessons Learned
Explain how this project has shaped your understanding of statistics and data analysis.

# Conclusion

Summarize your reflections and discuss how the skills and knowledge gained will benefit you in future coursework or projects.

# References
If applicable, include any citations for external data sources, articles, or R data packages.


# Appendix
Include any additional graphs, tables, or analysis outputs here if necessary. The length of the report should not be more than 3 pages.

