---
title: "R Commands for Applied Statistics"
subtitle: "Math 260"
format: 
  html:
    toc: true
    toc-depth: 2
    toc-title: "Jump to"
    theme: flatly
    embed-resources: true
    code-copy: true
execute:
  warning: false
  message: false
editor: visual
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE)
options(digits = 4)
library(mosaic)
library(Lock5Data)
set.seed(123)
```

```{=html}
<div id="filter-bar">
  <input type="text" id="cmd-filter" placeholder="Type to search — try histogram, average, spread, boxplot…" autocomplete="off" spellcheck="false">
  <span id="filter-count"></span>
</div>
<p id="filter-empty">Nothing matches that. Try a shorter word, or clear the box to see everything.</p>
<style>
:root {
  --k-green: #5A7A5E;
  --k-green-dark: #46603F;
  --k-tint: #E8F0E9;
}
h1.title { color: var(--k-green); }
.subtitle { color: #7b8a8b; font-weight: 400; }
#title-block-header { 
  border-bottom: 3px solid var(--k-green); 
  padding-bottom: 0.5rem; margin-bottom: 0.5rem;
}
main h1 {
  color: var(--k-green); font-size: 1.5rem;
  border-bottom: 2px solid var(--k-green);
  padding-bottom: 0.2rem; margin-top: 2.5rem;
}
main h2 { color: var(--k-green-dark); font-size: 1.15rem; margin-top: 1.6rem; }
div.sourceCode { border-left: 3px solid var(--k-green); }
.callout-note { 
  background: var(--k-tint) !important; 
  border-left-color: var(--k-green) !important; 
}
#filter-bar {
  position: sticky; top: 0; z-index: 50;
  background: #fff; padding: 0.75rem 0;
  border-bottom: 2px solid var(--k-green); margin-bottom: 1.5rem;
  display: flex; align-items: center; gap: 0.75rem;
}
#cmd-filter {
  flex: 1; font-size: 1rem; padding: 0.55rem 0.8rem;
  border: 1px solid #b7bfc6; border-radius: 4px;
  font-family: inherit; color: #2c3e50;
}
#cmd-filter:focus {
  outline: 2px solid var(--k-green); outline-offset: 1px; border-color: transparent;
}
#filter-count { font-size: 0.85rem; color: var(--k-green-dark); white-space: nowrap; }
#filter-empty { display: none; color: #7b8a8b; font-style: italic; }
.kw { display: none; }
@media print {
  #filter-bar, #filter-empty, #TOC, .toc-active { display: none !important; }
  section.level2 { break-inside: avoid; }
}
</style>
```

::: callout-note
## How to read a command

Almost everything below follows one pattern:

```         
goal(y ~ x, data = mydata)
```

The `~` is read **"described by"** or **"depends on"**. When there is only one variable, leave the left side off: `~ x`.

For plots, the left side of `~` is the **vertical** axis and the right side is the **horizontal** axis. Remember that **x** is the e**X**planatory variable, so it goes on the right.

Use `|>` to send a plot into another plotting command. You will see `%>%` in older code and in answers from AI assistants — it does the same thing, but use `|>` in this course.
:::

# Getting started

## Load the packages you need

Put this at the top of every Quarto file. Without it, R will not know what `favstats()` or `gf_histogram()` mean.

```{r}
#| eval: false
library(mosaic)
library(Lock5Data)
```

[packages, library, load, mosaic, ggformula, start]{.kw}

## Get help on a command

```{r}
#| eval: false
?favstats
help(gf_histogram)
```

[help, documentation, arguments, options, stuck]{.kw}

## See how big a dataset is

```{r}
dim(StudentSurvey)
```

The first number is rows (cases), the second is columns (variables). Use `nrow()` if you want only the row count.

```{r}
nrow(StudentSurvey)
```

[dimensions, size, how many rows, how many columns, cases, dim, nrow]{.kw}

## See what is in a dataset

```{r}
names(StudentSurvey)
```

```{r}
head(StudentSurvey)
```

`View(StudentSurvey)` opens a spreadsheet view. Type it at the console — do not put it in a code chunk.

[variables, columns, look at data, dataset, view, head, names]{.kw}

# One categorical variable

## Count how many are in each category

```{r}
tally(~ Award, data = StudentSurvey)
```

[table, frequency, counts, how many]{.kw}

## Show percents or proportions instead of counts

```{r}
tally(~ Award, data = StudentSurvey, 
      format = "percent")
```

```{r}
tally(~ Award, data = StudentSurvey, 
      format = "proportion")
```

Add `margins = TRUE` to include a total row.

[percent, proportion, relative frequency, share]{.kw}

## Bar chart of counts

```{r}
gf_bar(~ Award, data = StudentSurvey)
```

[bar graph, bar chart, categorical plot]{.kw}

## Bar chart of proportions

```{r}
gf_props(~ Award, data = StudentSurvey, 
         fill = "forestgreen")
```

Use `gf_props()` for proportions rather than modifying `gf_bar()`.

[proportion bar chart, percent bar graph]{.kw}

## Turn a bar chart sideways

```{r}
gf_bar(~ Award, data = StudentSurvey) |>
  gf_refine(coord_flip())
```

Long category names are much easier to read this way.

[horizontal, flip, sideways, rotate, coord_flip, long labels]{.kw}

# Two categorical variables

## Two-way table

```{r}
tally(~ Award + Smoke, data = StudentSurvey)
```

[cross tabulation, contingency table, two way, crosstab]{.kw}

## Conditional proportions

The `|` means "for each value of". This gives proportions **within** each smoking group.

```{r}
tally(~ Award | Smoke, data = StudentSurvey, 
      format = "proportion")
```

[conditional, given, within group, row percent, column percent]{.kw}

## Grouped bar chart

```{r}
gf_props(~ Sex, fill = ~ Award, 
         position = "dodge", 
         data = StudentSurvey)
```

[side by side, grouped bars, compare groups, dodge]{.kw}

## Segmented bar chart

```{r}
gf_props(~ Sex, fill = ~ Award, 
         position = "fill", 
         data = StudentSurvey)
```

`position = "fill"` makes every bar the same height so you compare proportions rather than counts. Use `"stack"` if you want the raw counts stacked instead.

[stacked, segmented, fill, stack, proportions by group]{.kw}

# One quantitative variable

## All the summary statistics at once

```{r}
favstats(~ Height, data = StudentSurvey)
```

This gives the minimum, quartiles, median, maximum, mean, standard deviation, sample size, and number of missing values. The IQR is not shown — subtract Q1 from Q3.

[mean, median, average, standard deviation, sd, quartiles, five number summary, min, max, center, spread, typical, IQR, missing]{.kw}

## One statistic at a time

```{r}
#| eval: false
mean(~ Height, data = StudentSurvey)
median(~ Height, data = StudentSurvey)
sd(~ Height, data = StudentSurvey)
IQR(~ Height, data = StudentSurvey)
quantile(~ Height, data = StudentSurvey)
```

[mean, median, sd, IQR, quantile, percentile, single statistic]{.kw}

## Histogram

```{r}
gf_histogram(~ Height, data = StudentSurvey)
```

[shape, distribution, skew, symmetric, plot one variable]{.kw}

## Histogram with a bin width you choose

```{r}
gf_histogram(~ Height, data = StudentSurvey, 
             binwidth = 2)
```

R picks a bin width if you do not. Changing it can change what the shape appears to be, so it is worth trying more than one.

[binwidth, bins, bars, width, resolution]{.kw}

## Dotplot

```{r}
gf_dotplot(~ Height, data = StudentSurvey, 
           dotsize = 0.25)
```

Every dot is one observation, which makes this a good choice for small datasets. Adjust `dotsize` until the dots look reasonable — the default is often far too large.

[dot plot, small data, individual values, dotsize]{.kw}

## Boxplot of one variable

```{r}
gf_boxplot(~ Height, data = StudentSurvey) |>
  gf_refine(coord_flip())
```

[box plot, five number summary, outliers, whiskers, horizontal]{.kw}

# Comparing groups

## Summary statistics for each group

```{r}
favstats(TV ~ Sex, data = StudentSurvey)
```

Read `TV ~ Sex` as "TV hours described by sex".

[by group, compare, split, grouped statistics, two groups]{.kw}

## Side-by-side boxplots (horizontal)

```{r}
gf_boxplot(Sex ~ TV, data = StudentSurvey)
```

**Why the categorical variable comes first here.** The left side of `~` is always the vertical axis. In a horizontal boxplot the groups run down the side, so the categorical variable is the vertical-axis variable and goes on the left. Horizontal is the better default: group labels stay readable, and comparing centers left-to-right is easier than comparing them up-and-down.

Note that this is the opposite order from `favstats(TV ~ Sex, ...)`, where the formula describes a relationship rather than a pair of axes.

[compare groups, box plot by group, horizontal, side by side, which variable first]{.kw}

## Stacked histograms for comparing groups

```{r}
gf_histogram(~ Height | Sex ~ ., data = StudentSurvey, 
             binwidth = 2)
```

The `| Sex ~ .` puts each group in its own panel, stacked **vertically**. Stacking is what makes the comparison work: the panels share one horizontal scale, so centers and spreads line up directly above one another. Panels placed side by side do not let you do that.

**Why the dot?** The part after `|` is a small formula describing the *layout* of the panels — read it as **rows \~ columns**. Putting `Sex` on the left makes one row per group, so the panels stack. The `.` is a placeholder meaning "nothing here": there is no second variable splitting the columns, but the layout formula still needs both sides. Writing `. ~ Sex` would arrange the groups side by side in columns instead, and plain `| Sex` lets R choose the arrangement for you.

Setting `binwidth` matters here too — without it the panels may not use the same bins.

[faceted, panels, separate plots, vertical, stacked, by group histogram, conditioning, compare distributions, dot, layout, rows, columns]{.kw}

# Two quantitative variables

## Scatterplot

```{r}
gf_point(Weight ~ Height, data = StudentSurvey)
```

Read `Weight ~ Height` as "weight as a function of height". The response goes on the left; **x** is the e**X**planatory variable and goes on the right.

[scatter plot, association, relationship, x y plot, explanatory, response]{.kw}

## Scatterplot with a regression line

```{r}
gf_point(Weight ~ Height, data = StudentSurvey) |>
  gf_smooth(method = "lm")
```

Use `gf_smooth(method = "lm")` for a straight line. Leaving out `method` gives a curve that follows the data instead.

[regression line, line of best fit, trend line, add line, lm]{.kw}

## Color the points by a categorical variable

```{r}
gf_point(Weight ~ Height, color = ~ Sex, 
         data = StudentSurvey)
```

Note the `~` in front of `Sex`. Without it, R looks for a color named "Sex".

[color by group, third variable, groups on scatterplot]{.kw}

## Correlation

```{r}
cor(Weight ~ Height, data = StudentSurvey, 
    use = "complete.obs")
```

Without `use = "complete.obs"` this returns `NA`, because some students did not report both height and weight. That argument tells R to use only the cases that have both values. Check the `missing` column of `favstats()` if you are not sure whether a variable has missing values.

[correlation, r, strength, linear association, NA, missing, complete.obs]{.kw}

# Making a plot presentable

## Label the axes and add a title

```{r}
gf_point(Weight ~ Height, data = StudentSurvey) |>
  gf_labs(title = "Weight versus height",
          x = "Height (inches)",
          y = "Weight (pounds)")
```

This works with any `gf_` plot. Variable names are rarely good axis labels — this is where units belong. Anything you hand in or put on a poster should have real labels.

[axis labels, title, units, caption, gf_labs, label axes, presentable]{.kw}

## Label one or two particular points

Sometimes one point is obviously unusual and you want to say what it is — the classic example is the Palm Beach County point on a plot of Florida's 2000 presidential vote. Read the point's position off the plot, then place text beside it:

```{r}
gf_point(Weight ~ Height, data = StudentSurvey) |>
  gf_refine(annotate("text", 
                     x = 78, y = 220, 
                     label = "unusually tall"))
```

`x` and `y` are the coordinates where the **text** goes, in the units of your axes. Start with the point's own position, then shift one of them until the label sits beside the point instead of on top of it.

`annotate()` comes from ggplot2, so it has to be wrapped in `gf_refine()` to work inside a `|>` chain — the same wrapper used for `coord_flip()`.

To draw a circle around the point as well, add a second annotation:

```{r}
#| eval: false
gf_point(Weight ~ Height, data = StudentSurvey) |>
  gf_refine(annotate("text", 
                     x = 78, y = 220, 
                     label = "unusually tall"),
            annotate("point", 
                     x = 79, y = 200, 
                     shape = 1, size = 6, color = "red"))
```

[label a point, annotate, text on plot, name a point, identify, which point is that, arrow, circle, mark, palm beach]{.kw}

# Regression

## Fit a model and see the coefficients

```{r}
mymodel <- lm(Weight ~ Height, data = StudentSurvey)
mymodel
```

Printing the model object by name gives you the intercept and slope. That is all you need in this course.

[lm, linear model, slope, intercept, fit, equation]{.kw}

## Predicted values

```{r}
predict(mymodel)[1:5]
```

The `[1:5]` shows just the first five. Leave it off to see all of them.

[prediction, fitted values, y hat, predicted]{.kw}

## Residuals

```{r}
resid(mymodel)[1:5]
```

[residual, error, leftover, observed minus predicted]{.kw}

# Collecting data

## Take a random sample

```{r}
#| eval: false
sample(StudentSurvey, 25)
```

[random sample, sampling, select cases, survey]{.kw}

## Randomly assign subjects to treatment groups

First shuffle the roster into a random order:

```{r}
myclass <- data.frame(
  student = c("Ava", "Ben", "Cara", 
              "Dan", "Eve", "Finn", 
              "Gia", "Hal", "Ivy")
)

newlist <- shuffle(myclass)
head(newlist)
```

Then take the first third, the second third, and so on:

```{r}
# Group 1
slice(newlist, 1:3)
```

```{r}
# Group 2
slice(newlist, 4:6)
```

```{r}
# Group 3
slice(newlist, 7:9)
```

The shuffle is what makes the assignment random — the slicing afterward is just bookkeeping. Putting `set.seed(123)` before `shuffle()` makes the result repeatable, which is useful when you want the same groups every time you render.

[random assignment, experiment, treatment groups, shuffle, slice, randomize, set.seed, experimental design]{.kw}

## Send data to StatKey

StatKey needs a CSV file. This writes one into your project folder, which you then upload.

```{r}
#| eval: false
write.csv(StudentSurvey, "mydata.csv", 
          row.names = FALSE)
```

[statkey, csv, export, upload, bootstrap, randomization test, save data]{.kw}

# Inference

## Confidence interval and test for one mean

```{r}
t.test(~ Height, data = StudentSurvey)
```

[t test, confidence interval, one mean, CI, p-value]{.kw}

## Comparing two means

```{r}
t.test(TV ~ Sex, data = StudentSurvey)
```

If you have subsetted the data first with `filter()`, add `droplevels()` before running this or R may complain about groups that are no longer there.

[two sample t test, difference in means, compare two groups, droplevels]{.kw}

# Working with your data

## Keep only some rows

```{r}
smokers <- filter(StudentSurvey, Smoke == "Yes")
nrow(smokers)
```

Use `==` for "is equal to", not a single `=`.

[filter, subset, select rows, only, exclude, condition]{.kw}

## Create a new variable

```{r}
#| eval: false
StudentSurvey <- mutate(StudentSurvey, 
                        TVperDay = TV / 7)
```

[mutate, new column, compute, transform, add variable]{.kw}

## Basic calculation

R works as a calculator at the console: `+ - * / ^ ( )` along with `log()`, `exp()`, and `sqrt()`.

[calculator, arithmetic, square root, log, exponent]{.kw}

# When something goes wrong

## "could not find function"

You did not load the package. Run `library(mosaic)` and try again. In a Quarto file, that line has to be in a chunk **above** the one you are running.

[error, could not find function, package not loaded, library]{.kw}

## "object 'Height' not found"

You left out `data =`. R looked for a loose variable named `Height` instead of a column in your dataset.

[error, object not found, data argument, variable name]{.kw}

## "unexpected symbol" or "unexpected ')'"

Usually a missing comma between arguments, or an unmatched parenthesis or quotation mark.

[error, syntax, unexpected, comma, parenthesis, typo]{.kw}

## The answer came out as `NA`

Your variable has missing values, and the command refused to guess what to do about them. For `cor()`, add `use = "complete.obs"`. For `mean()` and similar, add `na.rm = TRUE`.

[NA, missing, not available, na.rm, complete.obs, no answer]{.kw}

## The console shows `+` and will not respond

R is waiting for you to finish a command. Press **Esc** to cancel and start over.

[stuck, plus sign, console, hanging, escape]{.kw}

## No plot appeared

Check whether you saved it to a name — `p <- gf_histogram(...)` stores the plot instead of showing it. Type `p` to display it.

[plot missing, nothing happened, no output, blank]{.kw}

## A warning about rows being removed

Your variable has missing values. That is usually fine — it just means those cases were left out of the plot.

[warning, removed rows, missing values, incomplete cases]{.kw}

```{=html}
<script>
(function () {
  var box = document.getElementById('cmd-filter');
  var count = document.getElementById('filter-count');
  var empty = document.getElementById('filter-empty');
  var entries = Array.prototype.slice.call(document.querySelectorAll('section.level2'));
  var groups = Array.prototype.slice.call(document.querySelectorAll('section.level1'));
  var toc = document.getElementById('TOC');

  function apply() {
    var q = box.value.toLowerCase().trim();
    var shown = 0;

    entries.forEach(function (e) {
      var hit = q === '' || e.textContent.toLowerCase().indexOf(q) !== -1;
      e.style.display = hit ? '' : 'none';
      if (hit) shown++;
    });

    groups.forEach(function (g) {
      var kids = Array.prototype.slice.call(g.querySelectorAll('section.level2'));
      var any = kids.length === 0
        ? (q === '' || g.textContent.toLowerCase().indexOf(q) !== -1)
        : kids.some(function (k) { return k.style.display !== 'none'; });
      g.style.display = any ? '' : 'none';
    });

    if (q === '') {
      count.textContent = '';
      empty.style.display = 'none';
      if (toc) toc.style.display = '';
    } else {
      count.textContent = shown + (shown === 1 ? ' match' : ' matches');
      empty.style.display = shown === 0 ? 'block' : 'none';
      if (toc) toc.style.display = 'none';
    }
  }

  box.addEventListener('input', apply);

  box.addEventListener('keydown', function (e) {
    if (e.key === 'Escape') { box.value = ''; apply(); box.blur(); }
  });

  document.addEventListener('keydown', function (e) {
    if (e.key === '/' && document.activeElement !== box) {
      e.preventDefault();
      box.focus();
    }
  });
})();
</script>
```
