---
title: "2.4 Outliers, Boxplots, and Quantitative/Categorical Relationships"
subtitle: "Math 261"
author: "Eric Nordmoe"
format:
  revealjs:
    css: custom.css
    chalkboard: true
    theme: default
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE)
```

```{r, echo=F,message=FALSE}
library(mosaic)
library(Lock5Data)
library(palmerpenguins)
```

## Agenda


* One more visual display: the boxplot

* Investigating relationships between quantitative and categorical relationships
   + Comparing quantitative variables across groups
   


## Outliers

-   Often best identified using judgement by inspection of a visual data display.

-   **Rule of Thumb** suspected outliers are data values more than 1.5 IQRs beyond the quartiles

-   A data value is a **suspected outlier** if it is

    -   Smaller than the **Lower Fence**= $Q_1$ - 1.5 (IQR)

    -   Greater than the **Upper Fence**= $Q_3$ + 1.5 (IQR)


## Boxplot

-   A **boxplot** is

    -   A visual depiction of the 5-number summary.

    -   Especially useful for a rough comparison of the center and spread of a variable across several sub-groups.

## Constructing a Boxplot

1)  Draw a numerical scale appropriate for the data.

2)  Draw a box stretching from $Q_1$ to $Q_3$ representing the middle 50% of the data.

3)  Divide the box with a line at the median.

4)  Draw a line (whisker) from each quartile to the most extreme data value that is **not an outlier**.

5)  Identify each outlier individually by plotting with a symbol such as an asterisk or dot.

**Note:** Use side-by-side boxes to compare sub-groups.


## Boxplot example comparing groups: Wrong Way

```{r,fig.height=3.5}
gf_boxplot(Calories ~ PriorSmoke, data = NutritionStudy)
```

## Boxplot example comparing groups: Right Way

```{r,fig.height=3.5}
gf_boxplot(Calories ~ factor(PriorSmoke), data = NutritionStudy)
```

## Advanced Boxplot for RStudio Masters

```{r,fig.height=2}
gf_boxplot(Calories ~ factor(PriorSmoke), data = NutritionStudy,
           outlier.shape = NA) %>%
  gf_jitter(width = .1, alpha = .4) %>%
  gf_refine(coord_flip())
```

-   Overlay points using `jitter` on top of boxes to show more detail
-   `alpha` used to make points transparent (between 0 and 1)

## Comparing Numerical Statistics

Use `favstats()` to compare numerical statistics for a *quantitative variable* across *categories* of a categorical variable:

```{r}
favstats(Calories ~ PriorSmoke, data = NutritionStudy)
```

## Quantitative and Categorical Relationships

-   Compare the distribution of a quantitative variable across groups determined by the categorical variable(s).

    -   Use side-by-side boxplots for an overview.

    -   Use comparative histograms or dotplots to see more detail.

-   Any of the statistics we use for a quantitative variable can be looked at separately for each level of a categorical variable.

    -   Often interested in the **difference in means** $=\bar x_1- \bar x_2$
