---
title: "2.4 Boxplots and Quantitative/ Categorical Relationships"
author: "Math 260 Applied Statistics"
format:
  revealjs:
    chalkboard: true
    theme: [default, custom.scss]
    reveal-options:
      slideNumber: true
title-slide-attributes:
  data-background-color: "#EA6820"
from: markdown+emoji
execute:
  echo: true
  cache: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE,warning=FALSE)
library(mosaic)
library(Lock5Data)
```



## Outliers

* An outlier is an observed value that is notably distinct from the other values in a dataset.

* Often best identified using judgement by inspection of a visual data display.

* __Rule of Thumb__ suspected outliers are data values more than 1.5 IQRs beyond the quartiles

. . . 

* A data value is a __suspected outlier__ if it is

     + Smaller than the **Lower Fence**= $Q_1$ - 1.5 (IQR)
   
     + Greater than the **Upper Fence**= $Q_3$ + 1.5 (IQR)
     


## Boxplot

* A **boxplot** is 

    + A visual depiction of the 5-number summary.

    + Especially useful for a rough comparison of the center and spread of a variable across several sub-groups.


## Constructing a Boxplot 

1. Draw a numerical scale appropriate for the data.

2. Draw a box stretching from $Q_1$ to $Q_3$ representing the middle 50% of the data.

3. Divide the box with a line at the median.

4. Draw a line (whisker) from each quartile to the most extreme data value that is **not an outlier**.

5. Identify each outlier individually by plotting with a symbol such as an asterisk or dot.

**Note:** Use side-by-side boxes to compare sub-groups.



## Boxplot example comparing two groups

```{r,fig.height=5}
gf_boxplot(Calories ~ Smoke, data = NutritionStudy)
```



## Advanced Boxplot for RStudio Masters 

```{r,fig.height=2.5}
gf_boxplot(Smoke ~ Calories, data = NutritionStudy, 
           outlier.shape = NA) |>   
  gf_jitter(height = .1, alpha = .4) 
```
* Overlay points using `jitter` on top of boxes to show more detail
* `alpha` used to make points transparent (between 0 and 1)  
* `outlier.shape = NA` avoids plotting outlier point twice.



## Quantitative and Categorical Relationships

* Compare the distribution of a quantitative variable across groups determined by the categorical variable(s).

    + Use side-by-side boxplots for an overview.
   
    + Use comparative histograms or dotplots to see more detail.
    
. . . 

* Any of the statistics we use for a quantitative variable can be looked at separately for each level of a categorical variable.

    + Often interested in the **difference in means** $=\bar x_1- \bar x_2$
