---
title: "2.5 Two Quantitative Variables: Scatterplot and Correlation"
author: "E. Nordmoe"
date: "Math 260"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE)
```


```{r, echo=F,message=FALSE}
library(mosaic)
library(Lock5Data)
```


# Exploratory Data Analysis for Pairs of Quantitative Variables

* Visual/Graphical

    + Scatterplot
   
* Numerical

    + Correlation coefficient
   
    + Linear regression (Section 2.6)

---

# Scatterplot 

* A **scatterplot** is the graph of the relationship between two quantitative variables.

* Describe the *association* between two variables by commenting on:  
    + Form
        - linear, curvilinear, clusters, patternless
    + Direction (if applicable)
        - Positive or negative
    + Strength
        - Strong, moderate, weak
---

# Association

* A **positive** association means that values of one variable tend to be **higher** when values of the other variable are higher.

* A **negative** association means that values of one variable tend to be **lower** when values of the other variable are higher.

* Two variables are **not associated**  (independent) if knowing the value of one variable does not give you any information about the value of the other variable.

---

# Scatterplots with R: Example 1

```{r,fig.height=4.5}
gf_point(Alkalinity ~ pH,data = FloridaLakes, 
         xlab = "Acidity (pH)")
```

---

# Scatterplots with R: Example 2

* Smoother useful for identifying nonlinear trends
```{r,fig.height=4.5}
gf_point(Alkalinity ~ pH,data = FloridaLakes,
         xlab = "Acidity (pH)") %>%
  gf_smooth()
```

---

# Scatterplots with R: Example 3

* Identify subgroups based on categorical variable using `color` or `shape`  
```{r,fig.height=4.5}
gf_point(WeekendSleep ~ WeekdaySleep, 
         color = ~ factor(EarlyClass), data = SleepStudy)
```

---

# Scatterplots with R: Example 4

* Extreme points can be labeled and omitted if need be.  
* `hjust=1` means right-justified, `hjust=0` means left-justified.

```{r,fig.height=4}
gf_point(Alcohol ~ Calories, data = NutritionStudy) +
        annotate("text", x = 6600, y = 203, hjust = 1,
                 label = "Case #62")
```

---

## Scatterplots with R: Example 4 (cont'd)

* Omit #62 and re-plot

```{r,fig.height=4.5}
gf_point(Alcohol ~ Calories, data = NutritionStudy[-c(62) , ])
```

---

## Scatterplots with R: Example 4 (cont'd)

* Create a subset first using the `filter()` function

```{r,fig.height=4.5}
NutritionStudy2 <- filter(NutritionStudy, Alcohol < 100)
gf_point(Alcohol ~ Calories, data = NutritionStudy2)
```

---

# Correlation

The **correlation coefficient** is a numerical measure of the *strength and direction* of **linear** association between two quantitative variables.

* Sample correlation coefficient: $r$
* Population correlation coefficient: $\rho$

---

## Properties of the Correlation Coefficient

1. $-1\le r\le 1$

2. The sign of $r$ indicates the direction of association:
    + positive linear association: $r>0$
    + negative linear association: $r<0$
    + no linear assocation: $r=0$
   
3. The closer $r$ is to $\pm 1$, the stronger the linear association.

4. $r$ has no units and does not depend on the units of measurement. 

5. Swapping $X$ and $Y$ does not change the correlation coefficient.


---

# Using R to Compute Correlations

* Use the formula syntax:
```{r}
cor(Calories ~ Alcohol, data = NutritionStudy)
cor(Calories ~ Alcohol, data = NutritionStudy[-c(62), ])
```

---

# Correlation Cautions

1. $r$ is **not resistant**.
    + Correlation can be heavily affected by outliers. 
    + Always plot your data!
--

2. $r = 0$ means no **linear** association.  The variables could still be otherwise associated.  
    + Always plot your data!
--

3. Correlation does not imply causation!
