---
title: "2.3  One Quantitative Variable: Measures of Spread"
author: "E. Nordmoe"
date: "Math 260"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE)
```


```{r, echo=F,message=FALSE}
library(mosaic)
library(Lock5Data)
```



# Standard Deviation

The **standard deviation** for a quantitative variable measures the spread of the data.

$$
s=\sqrt{\frac{\displaystyle\sum_{i=1}^n (x_i-\bar x)^2}{n-1}}
$$

--

* Gives a rough estimate of the *typical distance* of data values from the mean.
   + Std Dev $\approx\sqrt{\mbox{average of (deviations from average)}^2}$

--

* The larger the standard deviation, the greater the spread.

--

* Has the same units as the original data.

---

# Intepreting the Standard Deviation: The 95% Rule 

* If a distribution of data is **approximately bell-shaped**, about 95% of the data should fall within two standard deviations of the mean.

--

* For a sample, about 95% of the data should be between $\bar x-2s$ and
$\bar x+2s$.

--

* Note: In practice, this rule-of-thumb works reasonably well even when the data are not bell-shaped.

---

# SleepStudy Data 

### Description

* Data from a study of sleep patterns for college students.

### Format

* A dataset with 253 observations on  27 variables.

---

# Illustration of 95% Rule

```{r,echo=FALSE,fig.height=4}
xbar <- mean(~WeekdaySleep, data = SleepStudy)
st.dev <- sd(~WeekdaySleep, data = SleepStudy)
ggplot(SleepStudy, aes(x = WeekdaySleep)) +
        geom_histogram(fill = "white", color = "black", breaks = seq(xbar - 4 * st.dev, xbar + 4 * st.dev, st.dev)) +
        geom_histogram(data = subset(SleepStudy, WeekdaySleep >= xbar - 2 * st.dev & WeekdaySleep <= xbar + 2 * st.dev), breaks = seq(xbar - 4 * st.dev, xbar + 4 * st.dev, st.dev), color = "black", fill = "red", alpha =.5)
```

--

* Despite skewness, about 95% of data lie between 7.87-2(1.17) and 7.87+2(1.17).  

--

* Note: It's actually 246/253=.972 for these data.

---

# Percentile

The **p
$^\text{th}$ percentile**  is the value which is greather than **p%** of the data.

* A test score in the 80th percentile is greater than at least 80% of the scores on the test.

---

# Example: 80th Percentile 

* Illustrate using an idealized smoothed histogram:

```{r}
quantile( ~ WeekdaySleep, data = SleepStudy, probs = c(.8))
```
```{r,echo=FALSE,fig.height=3}
#  calculate quantiles
qp8 <- quantile(~WeekdaySleep,data=SleepStudy,probs=c(.8))
# generate the plot
x <- density(SleepStudy$WeekdaySleep, n = 253)$x
y <- density(SleepStudy$WeekdaySleep, n = 253)$y
gg <- data.frame(SleepStudy, x, y)
ggplot(SleepStudy) + stat_density(aes(x = WeekdaySleep), geom = "line") +
        geom_ribbon(data = subset(gg, x <= qp8),
                    aes(x = x, ymax = y), ymin = 0,fill = "red", alpha = 0.5) +
        geom_ribbon(data = subset(gg, x >= qp8),
                    aes(x = x, ymax = y), ymin = 0, fill = "blue", alpha = 0.5)+
        annotate("text", x = 7, y = 0.08, label = "0.80")+
        annotate("text", x = 9.2, y = 0.08, label = "0.20")
```

---

# The z-Score

The **
$z$-score** for a data value, $x$, from a sample mean mean 
$\bar x$ and standard deviation $s$ is defined to be

$$
z-\text{score}=\frac{x-\bar x}{s}
$$

* The $z$-score tells how many standard deviations the value is from the mean.

* **Note:** For a population, replace $\bar x$ by $\mu$ and $s$ by $\sigma$. 

---


# z-Scores

* A z-score puts values on a common scale

* Values farther from 0 are more extreme 

* For bell-shaped distributions, about 95% of all z-scores fall between -2 and 2

    + Identify extreme values with $|z|>2$.


???

Carry out example(s) using R.  Which score is higher?

---

# z-Scores: Your turn

* Is student 8 relatively more extreme on `DepressionScore` or `AnxietyScore`?  

```{r}
favstats(~ DepressionScore, data = SleepStudy)
favstats(~ AnxietyScore , data = SleepStudy)
SleepStudy[8 , c("DepressionScore", "AnxietyScore")]
```

---

# Other Measures of Location

* **Maximum** largest data value

* **Minimum** smallest data value

* Quartiles
     + $Q_1$ (lower quartile) is the 25th percentile, with about $1/4$ of the data below it 
   
     + $Q_3$ (upper quartile) is the 75th percentile, with about $1/4$ of the data above it
     
---
   
# Five Number Summary

**Five Number Summary** = (minimum, $Q_1$, median, $Q_3$, maximum)

* The five number summary divides the dataset into fourths
```{r}
favstats(~ WeekdaySleep, data = SleepStudy)
```

---

# Other Measures of Spread

**Range** = Max - Min

```{r}
diff(range(~ WeekdaySleep, data = SleepStudy))
```


**Interquartile Range (IQR)**= $Q_3-Q_1$.

```{r}
IQR(~ WeekdaySleep , data = SleepStudy)
```

---

# Summary

| Chateracteristic |   Resistant  |         Non-Resistant         |
|------------------|:------------:|:-----------------------------:|
| Center           |    Median    |              Mean             |
| Spread           |      IQR     | Standard deviation <br> Range |
| Location         | $Q_1$, $Q_3$ |            Min, Max           |

* Note that extreme percentiles (such as the 1st or 99th percentiles would be non-resistant unless the sample size were very large).

