---
title: "2.1 Categorical Variables"
subtitle: "Math 261"
author: "Eric Nordmoe"
format:
  revealjs:
    chalkboard: true
    theme: default
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE)
```

```{r, echo=F,message=FALSE}
library(mosaic)
library(Lock5Data)
library(palmerpenguins)
```

## Descriptive Statistics

-   To make sense of data, we need to **summarize and visualize** it.

-   **Descriptive statistics** summarizing and visualizing variables and relationships between variables.

-   Different methods are used for categorical and quantitative variables.

## Summarizing Single Categorical Variables {.smaller}

Summary statistics

-   **Frequency table** table showing number of cases in each category

-   **Proportion** number in category/total sample size

-   **Relative frequency table** table showing proportion of cases in each category

. . .

Visualizations

-   **Bar chart** height of the bar shows the number (or proportion or percent) of cases in each category

-   **Pie chart** relative area represents proportion in each category (less preferred)

## Frequency Table of Counts

```{r}
tally(~ LarkOwl, data = SleepStudy)
```

## Relative Frequency Table

```{r}
tally(~ LarkOwl, format = "proportion", margins = TRUE,
      data = SleepStudy)
```

## Bar Chart of Single Variable

```{r}
gf_bar(~ LarkOwl, data = SleepStudy)
```

## Summarizing Two Categorical Variables: Two-Way Tables

Frequency table used to look at the relationship between two categorical variables

-   Variable 1 categories: rows
-   Variable 2 categories: columns
-   Cell counts (frequencies): number of cases in corresponding row and column categories

## Two-Way Tables Example

Cross classify two categorical variables to investigate their relationship:

```{r}
tally(~ EarlyClass + LarkOwl, margins = TRUE,
      data = SleepStudy)
```

::: incremental
-   What proportion of students have early classes and are Larks?
-   What proportion of students with early classes are Larks?
-   What proportion of Larks have early classes?
:::

## Two-Way Tables Example with Proportions

```{r}
tally(~ EarlyClass + LarkOwl, margins = TRUE,
      format = "proportion", data = SleepStudy)
```

## Two-Way Tables Example with Conditional Proportions: Version 1 {.smaller}

To compare Larks and Owls:

```{r}
tally(EarlyClass ~ LarkOwl, margins = TRUE,
      format = "proportion", data = SleepStudy)
```



## Two-Way Tables Example with Conditional Proportions: Version 2 {.smaller}

To compare students with and without early classes:


```{r}
tally(LarkOwl ~ EarlyClass, margins = TRUE,
      format = "proportion", data = SleepStudy)
```

## Segmented Bar Charts for Two Variables: The Wrong Way {.smaller}

-   **Segmented bar chart** graphical display of a two-way table of counts

```{r,fig.height = 3}
gf_bar(~ LarkOwl, fill = ~ EarlyClass, data = SleepStudy)
```

-  Question: Why didn't `fill = ~ EarlyClass` have any effect?

. . . 

-  Answer: It's coded as a 0-1 variable.

## Segmented Bar Charts for Two Variables: The Right Way {.smaller}

-   **Segmented bar chart** graphical display of a two-way table of counts

Note: We use `factor()` to make R treat EarlyClass as a categorical variable.

```{r,fig.height = 2.5}
gf_bar(~ LarkOwl,fill = ~ factor(EarlyClass), data = SleepStudy)
```

## Comparative Stacked Bar Charts Comparing Proportions

```{r,fig.height = 3.5}
gf_props(~ LarkOwl, data = SleepStudy,
         fill = ~ factor(EarlyClass), position = "stack")
```

## Side-by-Side Bar Charts

-   **Side-by-side** height of each bar is the number of the corresponding cell in the two-way table

```{r, fig.height = 3}
gf_bar(~ LarkOwl, data = SleepStudy,
       fill = ~ factor(EarlyClass), position = "dodge")
```

## Side-by-Side Bar Charts Comparing Proportions

-   **Side-by-side** height of each bar is now 1 to allow comparison of proportions within each LarkOwl group:

```{r, fig.height = 3}
gf_props(~ LarkOwl, data = SleepStudy,
       fill = ~ factor(EarlyClass), position = "fill")
```

## Summary  {.smaller}

1. **Frequency Tables and Proportions**: Summarize single categorical variables using frequency tables to count cases in each category, and calculate proportions to understand the relative size of each category.

2. **Two-Way Tables**: Explore relationships between two categorical variables using two-way tables, which cross-classify variables to investigate how they are related and to calculate conditional proportions to compare categories.

3. **Bar Charts for Visualization**: Visualize categorical data with bar charts, including segmented bar charts to display two-way table counts, and side-by-side bar charts to compare proportions within categories.