---
title: "2.6 Two Quantitative Variables: Linear Regression"
author: "E. Nordmoe"
date: "Math 260"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE)
```


```{r, echo=F,message=FALSE}
library(mosaic)
library(Lock5Data)
```





# The Regression Line


* The process of fitting a line to a set of data is called **linear regression** and the line of best fit is called the **regression line**. 

* The estimated regression line is
$$
\hat y = a +bx
$$
where $x$ is the quantitative explanatory variable and $\hat y$ is the predicted quantitative response.

* The regression equation can be used to predict $y$ for a given value of $x$.

---

# Finding the Regression Line

* The **observed response variable**, $y$, is the response value for a particular data point.

* The **predicted response variable**, $\hat y$, is the reponse value that would be predicted for a given $x$ value based on a model.

* The best fitting line is the one which makes the predicted values closest to the actual values.

---

## Residuals

* The **residual** for each data point is 
$$
\text{observed}-\text{predicted}=y-\hat y
$$

* The residual is also the *vertical* distance from each point to the fitted line.

---

# Least Squares Line

* The **least squares line** is the line which minimizes the sum of **squared** residuals:

$$
\text{sum of (residuals}^2)=\sum_{i=1}^n (y_i-\hat y_i)^2
$$

* Try the StatsPortal [correlation and regression applet](http://digitalfirst.bfwpub.com/stats_applet/stats_applet_5_correg.html)

* For real data, use R to find the equation of the line.

**Note**: The "regression line" and the "least squares line" are the same thing.

---

# Slope and Intercept Interpretations

The estimated regression line is 
$$
\hat y = a + bx
$$

* **Slope (_b_)**: increase in  predicted $y$ for every 1 unit increase in $x$.

* **Intercept (_a_)**: predicted $y$ value when $x=0$

Note: In practice, interpretation of the slope is typically most important.

---

# Regression Cautions 

1. **Do not extrapolate!** Do not use the regression equation or line to predict outside the range of $x$ values available in your data.

2. Only use the regession line if the association is **approximately linear**. **Always plot your data!** 

3. **Beware of outliers!** Outliers (especially  values that are extreme on the $x$ axis) can be very influential on the regression line. 

4. **Correlation is not causation $\ldots$ and neither is regression!** Higher values of $x$ may correspond to higher (or lower) predicted values of $y$, but this does **not** mean that changing $x$ will *cause* $y$ to increase or decrease.

