---
title: "2.5 Two Quantitative Variables: Scatterplot and Correlation"
author: "Math 260 Applied Statistics"
format:
  revealjs:
    chalkboard: true
    theme: [default, custom.scss]
    reveal-options:
      slideNumber: true
title-slide-attributes:
  data-background-color: "#EA6820"
from: markdown+emoji
execute:
  echo: true
  cache: true
---

```{r, echo=F,message=FALSE}
library(mosaic)
library(Lock5Data)
```


## Exploratory Data Analysis for Pairs of Quantitative Variables

* Visual/Graphical

    + Scatterplot
   
* Numerical

    + Correlation coefficient
   
    + Linear regression (Section 2.6)


## Scatterplot  {.smaller}

* A **scatterplot** is the graph of the relationship between two quantitative variables.

* Describe the *association* between two variables by commenting on:  
    + Form
        - linear, curvilinear, clusters, patternless
    + Direction (if applicable)
        - Positive or negative
    + Strength
        - Strong, moderate, weak


## Association

* A **positive** association means that values of one variable tend to be **higher** when values of the other variable are higher.

* A **negative** association means that values of one variable tend to be **lower** when values of the other variable are higher.

* Two variables are **not associated**  (independent) if knowing the value of one variable does not give you any information about the value of the other variable.


## Scatterplots with R: Example 1

```{r}
gf_point(Alkalinity ~ pH,data = FloridaLakes, 
         xlab = "Acidity (pH)")
```


## Scatterplots with R: Example 2 {.smaller}

* Smoother useful for identifying nonlinear trends
```{r}
gf_point(Alkalinity ~ pH,data = FloridaLakes,
         xlab = "Acidity (pH)") |> 
  gf_smooth()
```


## Scatterplots with R: Example 3 {.smaller}

* Identify subgroups based on categorical variable using `color` or `shape`  
```{r}
gf_point(WeekendSleep ~ WeekdaySleep, 
         color = ~ factor(EarlyClass), data = SleepStudy)
```



## Scatterplots with R: Example 4 {.smaller}

* Extreme points can be labeled and omitted if need be.  
* `hjust=1` means right-justified, `hjust=0` means left-justified.

```{r}
gf_point(Alcohol ~ Calories, data = NutritionStudy) +
        annotate("text", x = 6600, y = 203, hjust = 1,
                 label = "Case #62")
```


## Scatterplots with R: Example 4 (cont'd) {.smaller}

* Omit the outlier using the `filter()` function to include only cases with `Alcohol` less than 100.

```{r}
NutritionStudy2 <- filter(NutritionStudy, Alcohol < 100)
gf_point(Alcohol ~ Calories, data = NutritionStudy2)
```


## Correlation

The **correlation coefficient** is a numerical measure of the *strength and direction* of **linear** association between two quantitative variables.

* Sample correlation coefficient: $r$
* Population correlation coefficient: $\rho$


## Properties of the Correlation Coefficient {.smaller}

1. $-1\le r\le 1$

2. The sign of $r$ indicates the direction of association:
    + positive linear association: $r>0$
    + negative linear association: $r<0$
    + no linear assocation: $r=0$
   
3. The closer $r$ is to $\pm 1$, the stronger the linear association.

4. The coefficient $r$ has no units and its value does not depend on the units of measurement. 

5. Interchanging $X$ and $Y$ does not change the correlation coefficient.



## Using R to Compute Correlations

* Use the formula syntax:
```{r}
cor(Calories ~ Alcohol, data = NutritionStudy)
cor(Calories ~ Alcohol, data = filter(NutritionStudy, Alcohol < 100))
```


## Correlation Cautions

1. $r$ is **not resistant**.
    + Correlation can be heavily affected by outliers. 
    + Always plot your data!

. . . 


2. $r = 0$ means no **linear** association.  The variables could still be otherwise associated.  
    + Always plot your data!
    
. . . 


3. Correlation does not imply causation!

## Summary {.smaller}

- **Scatterplots** visualize relationships between two quantitative variables, focusing on form, direction, and strength of association.

- **Association Types**
  - Positive: Both variables increase together.
  - Negative: One variable increases as the other decreases.
  - None: No discernible pattern between variables.

- **Correlation Coefficient ($r$)**
  - Measures strength and direction of linear relationships.
  - Ranges from -1 to 1: closer to $\pm 1$ indicates a stronger linear relationship.
  - Not resistant to outliers and should always be accompanied by a scatterplot for context.

- **Cautions**
  - Always visualize data to check for patterns not captured by correlation.
  - Remember that correlation does not imply causation.
