---
title: "2.4 Boxplots and Quantitative/ Categorical Relationships"
author: "Math 260 Applied Statistics"
format:
  revealjs:
    chalkboard: true
    theme: [default, custom.scss]
    reveal-options:
      slideNumber: true
title-slide-attributes:
  data-background-color: "#EA6820"
from: markdown+emoji
execute:
  echo: true
  cache: true
---

```{r, echo=F,message=FALSE}
library(mosaic)
library(Lock5Data)
```

## Agenda

- **Introduction to Boxplots**  
  - Definition and components of a boxplot
  - Constructing boxplots: step-by-step 

- **Analyzing Quantitative and Categorical Relationships**  
  - Comparing distributions of quantitative variables across categorical groups
  - Using boxplots for visual comparison

- **Advanced Visualization**  
  - Faceted plots: dotplots and histograms for group comparisons
  - Overlaying data by jittering for more detail

## Boxplot

-   A **boxplot** is

    -   A visual depiction of the 5-number summary.

    -   Especially useful for a rough comparison of the center and spread of a variable across several sub-groups.

## Constructing a Boxplot

1)  Draw a numerical scale appropriate for the data.

2)  Draw a box stretching from $Q_1$ to $Q_3$ representing the middle 50% of the data.

3)  Divide the box with a line at the median.

4)  Draw a line (whisker) from each quartile to the most extreme data value that is **not an outlier**.

5)  Identify each outlier individually by plotting with a symbol such as an asterisk or dot.

**Note:** Use side-by-side boxes to compare sub-groups.

## Boxplot example comparing groups: Vertical

```{r}
gf_boxplot(Quetelet ~ Smoke, data = NutritionStudy)
```

## Boxplot example comparing groups: Horizontal

```{r}
gf_boxplot(Smoke ~ Quetelet, data = NutritionStudy)
```

## Outliers

-   Often best identified using judgement by inspection of a visual data display.

-   **Rule of Thumb** suspected outliers are data values more than 1.5 IQRs beyond the quartiles

-   A data value is a **suspected outlier** if it is

    -   Smaller than the **Lower Fence**= $Q_1$ - 1.5 (IQR)

    -   Greater than the **Upper Fence**= $Q_3$ + 1.5 (IQR)

#  {background-color="#FAD9C7"}

::: {#title-slide .center}
[Your turn: Investigate Quetelet (BMI) by Prior Smoking Status (PriorSmoke)]{style="font-size: 1.5em; font-weight: bold;"}
:::

## Advanced Boxplot for RStudio Masters {.smaller}

```{r,fig.height=3}
gf_boxplot(factor(PriorSmoke) ~ Quetelet, data = NutritionStudy,
           outlier.shape = NA, ylab = "Prior Smoking Status") %>%
  gf_jitter(height = .1, alpha = .4) 
```

-   Overlays points using `jitter` on top of boxes to show more detail
-   `alpha` used to make points transparent (between 0 and 1)
-   Adds y label using `ylab`

# Faceted plots for more detailed comparisons {background-color="#FAD9C7"}

## Side-by-side Dotplots

-   For small samples, we can get a detailed plot of hours exercised for for each `Year` (1 = First Years, ...)
-   Use the vertical bar (`|`) to "condition" on the group variable.
-   `Year ~ .` specifies `Year` as the explanatory variable.

```{r}
#| output-location: slide
gf_dotplot(~Exercise | Year ~ ., data = ExerciseHours, dotsize = 0.5)
```

::: aside
Note: We actually could specify two variables such as `Year ~ Sex` to get a matrix of plots. Here, the `.` serves as a placeholder.
:::

## Faceted Histograms

For larger samples, we get side by side histograms of BMI (`Quetelet`) for each level of the `PriorSmoke` variable.

```{r}
#| output-location: slide
gf_histogram(~Quetelet | PriorSmoke ~ ., data = NutritionStudy)
```

## Comparing Numerical Statistics

Use `favstats()` to compare numerical statistics for a *quantitative variable* across *categories* of a categorical variable:

```{r}
favstats(Calories ~ PriorSmoke, data = NutritionStudy)
```

## Investigating Quantitative and Categorical Relationships  {.smaller}

-   Compare the distribution of a quantitative variable across groups determined by the categorical variable(s).

    -   Use side-by-side boxplots for an overview.

    -   Use comparative histograms or dotplots to see more detail.

-   Any of the statistics we use for a quantitative variable can be looked at separately for each level of a categorical variable.

    -   Often interested in the **difference in means** $=\bar x_1- \bar x_2$

More [details and examples](https://heroic-horse-92e142.netlify.app/describing_quantitative_variables_m260) in the R Resources on Moodle


