---
title: '2.5 Two Quantitative Variables: Scatterplot and Correlation'
subtitle: "Math 261"
author: "Eric Nordmoe"
format:
  revealjs:
    css: custom.css
    chalkboard: true
    theme: default
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE)
```

```{r, echo=F,message=FALSE}
library(mosaic)
library(Lock5Data)
library(palmerpenguins)
```

## Exploratory Data Analysis for Pairs of Quantitative Variables

-   Visual/Graphical

    -   Scatterplot

-   Numerical

    -   Correlation coefficient

    -   Linear regression (Section 2.6)

## It's all about relationships

```{r}
gf_point(Gestation ~ Longevity, data = MammalLongevity)
```

::: incremental
-   Overall pattern?
-   Outliers
:::

## Scatterplot {.smaller}

-   A **scatterplot** is the graph of the relationship between two quantitative variables.

-   Describe the *association* between two variables by commenting on:

    -   Form
        -   linear, curvilinear, clusters, patternless
    -   Direction (if applicable)
        -   Positive or negative
    -   Strength
        -   Strong, moderate, weak

## Association

-   A **positive** association means that values of one variable tend to be **higher** when values of the other variable are higher.

-   A **negative** association means that values of one variable tend to be **lower** when values of the other variable are higher.

-   Two variables are **not associated** (independent) if knowing the value of one variable does not give you any information about the value of the other variable.

## Scatterplots with R: Example 1 {.smaller}

```{r}
gf_point(Alkalinity ~ pH,data = FloridaLakes) %>%
  gf_labs(x = "Acidity (pH)")
```

## Scatterplots with R: Example 2 {.smaller}

-   Smoother useful for identifying nonlinear trends\

```{r}
gf_point(Alkalinity ~ pH,data = FloridaLakes) %>%
  gf_smooth() %>%
  gf_labs(x = "Acidity (pH)") 
```

## Scatterplots with R: Example 3 {.smaller}

-   Identify subgroups based on categorical variable using `color` or `shape`

```{r}
gf_point(Obese ~ PhysicalActivity, data = USStates, 
         color = ~factor(Region))
```

## Scatterplots with R: Example 4 {.smaller}

-   Use size to show a quantitative variable

```{r}
gf_point(Alkalinity ~ pH, data = FloridaLakes,
         size = ~NumSamples)
```

## Scatterplots with R: Example 5 {.smaller}

-   Extreme points can be labeled and omitted if need be.
-   `hjust = 1` means right-justified, `hjust = 0` means left-justified.

```{r,fig.height = 3}
gf_point(Alcohol ~ Calories, data = NutritionStudy) +
        annotate("text", x = 6600, y = 203, hjust = 1, label = "Case #62")
```

## Scatterplots with R: Example 5 (cont'd) {.smaller}

-   Omit #62 and re-plot using the `slice()` function

```{r,fig.height = 3}
gf_point(Alcohol ~ Calories,data = slice(NutritionStudy, -62))
```

## Scatterplots with R: Example 5 (cont'd) {.smaller}

-   Or create a subset first using `filter()`

```{r,fig.height = 3}
NutritionStudy2 <- filter(NutritionStudy, Alcohol < 100)
gf_point(Alcohol ~ Calories, data = NutritionStudy2)
```

## Correlation

The **correlation coefficient** is a numerical measure of the *strength and direction* of **linear** association between two quantitative variables.

-   Sample correlation coefficient: $r$\
-   Population correlation coefficient: $\rho$

## Properties of the Correlation Coefficient {.smaller}

1.  $-1\le r\le 1$

2.  The sign of $r$ indicates the direction of association:

    -   positive linear association: $r>0$\
    -   negative linear association: $r<0$\
    -   no linear assocation: $r = 0$

3.  The closer $r$ is to $\pm 1$, the stronger the linear association.

4.  $r$ has no units and does not depend on the units of measurement.

5.  Swapping $X$ and $Y$ does not change the correlation coefficient.

## Using R to Compute Correlations

-   Use the formula syntax:

```{r}
cor(Calories ~ Alcohol, data = NutritionStudy)
cor(Calories ~ Alcohol, data = slice(NutritionStudy, -62)) 
```

## Correlation Cautions

1.  $r$ is **not resistant**.
    -   Correlation can be heavily affected by outliers.\
    -   Always plot your data!
2.  $r = 0$ means no **linear** association. The variables could still be otherwise associated.
    -   Always plot your data!
3.  Correlation does not imply causation!
