---
title: "2.1 Categorical Variables"
author: "E. Nordmoe"
date: "Math 260"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(mosaic)
library(Lock5Data)
```


```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_duo_accent(
  primary_color = "#1381B0",
  secondary_color = "#FF961C",
  inverse_header_color = "#FFFFFF"
)
```



# Descriptive Statistics

* To make sense of data, we need to **summarize and visualize** it.

* **Descriptive statistics** summarizing and visualizing variables and relationships between variables.

* Different methods are used for categorical and quantitative variables.

---
class:  center, middle, inverse


# Numerical and Visual Descriptions for Categorical Variables

---

# Summarizing Single Categorical Variables

Summary statistics

* **Frequency table** table showing number of cases in each category

* **Proportion** number in category/total sample size

* **Relative frequency table** table showing proportion of cases in each category

Visualizations

* **Bar chart** height of the bar shows the number (or proportion or percent) of cases in each category

* **Pie chart** relative area represents proportion in each category (less preferred)

---

# Frequency Table of Counts

```{r}
tally(~LarkOwl, data = SleepStudy)
```

---

## Relative Frequency Table

```{r}
tally(~LarkOwl, format = "proportion",
      margins = TRUE, data = SleepStudy)
```


---

## Bar Chart of Single Variable
```{r, fig.height = 6}
gf_bar(~LarkOwl, data = SleepStudy)
```

---

# Summarizing Two Categorical Variables: Two-Way Tables

Frequency table used to look at the relationship between two categorical variables

* Variable 1 categories: rows
* Variable 2 categories: columns
* Cell counts (frequencies): number of cases in corresponding row and column categories

---

# Two-Way Tables Example

Cross classify two categorical variables to investigate their relationship:

```{r}
tally(~EarlyClass + LarkOwl, margins = TRUE,
      data=SleepStudy)
```

Note: For *EarlyClass*, 0 = "No" and 1 = "Yes"
--

* What proportion of students with early classes are Larks?
* What proportion of students without early classes are Larks?
* What proportion of Owls have early classes?

---

# Two-Way Tables Example with Proportions

```{r}
tally(~EarlyClass + LarkOwl, margins = TRUE,
      format = "proportion", data = SleepStudy)
```

---

# Two-Way Tables Example with Conditional Proportions

```{r}
tally(~EarlyClass | LarkOwl, margins = TRUE,
      format = "proportion", data = SleepStudy)
```

$\Rightarrow$ Conditional proportions are useful for comparing groups but be sure to compute the proportions in the correct direction.

---

# Bar Charts for Single Variables

```{r, fig.height = 6}
gf_bar( ~ LarkOwl, data = SleepStudy)
```

---

# Segmented Bar Charts for Two Variables

* **Segmented bar chart** graphical display of a two-way table of counts  

  * Note use of `factor()` to treat *EarlyClass* as categorical.

```{r,fig.height=4}
gf_bar(~ LarkOwl, fill = ~factor(EarlyClass), data = SleepStudy)
```


---

# Comparative Stacked Bar Charts Comparing Proportions

```{r,fig.height=4}
gf_bar(~LarkOwl, fill = ~factor(EarlyClass), data = SleepStudy, 
       position = "fill")
```

* Note that the function `gf_props()` can also be used to plot proportions when working with categorical data.

---

# Side-by-Side Bar Charts

* **Side-by-side** height of each bar is the number of the corresponding cell in the two-way table

```{r, fig.height=4}
gf_bar(~LarkOwl, fill = ~factor(EarlyClass), 
       data = SleepStudy, position = "dodge")
```


---
# Aesthetic Mappings

* We control the appearance of graphs in R using  *aesthetics*, things we can see

* Examples of aesthetics include:  
    + position (i.e., on the x and y axes)  
    + color ("outside" color)  
    + fill ("inside" color)
    + shape (of points)  
    + linetype  
    + size  
    
* Each type of graph accepts only a subset of possible aesthetics.  

* Using commands to create graphs makes it easy to reproduce or update them later.




