---
title: "3.1 Sampling Distributions"
subtitle: "Math 261"
format:
  revealjs:
    css: custom.css
    chalkboard: true
    theme: default
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE)
```

```{r, echo=F,message=FALSE}
library(mosaic)
library(Lock5Data)
```

## Statistical Inference

**Statistical inference** is the process of drawing conclusions about the entire *population* based on information in a *sample*.

. . .

A **parameter** is a number that describes some aspect of a **population.**

. . .

A **statistic** is a number that is computed from data in a **sample.**

::: notes
Re-draw diagram with pop and sample showing idea of sampling distribution and inference
:::

## Parameter vs Statistic

| Measure     | **Parameter** | **Statistic** |
|-------------|:-------------:|:-------------:|
| Mean        |     $\mu$     |   $\bar x$    |
| Proportion  |      $p$      |   $\hat p$    |
| Std Dev     |   $\sigma$    |      $s$      |
| Correlation |    $\rho$     |      $r$      |

-   We use a *statistic* from a *sample* as a **point estimate** of the corresponding *population parameter*.

## Important Points

-   Sample statistics vary from sample to sample.

-   **Key Questions** For a given sample statistic,

    -   What are plausible values for the population parameter?

    -   How much uncertainty is associated with the sample statistic?

-   **Key Answer** It depends on how much the statistic varies *from sample to sample*!

## Sampling Distribution

-   A **sampling distribution** is the distribution of sample statistics computed for different samples of the same size from the same population.

    -   A sampling distribution shows us how the sample statistic varies *from sample to sample*.

## Center and Shape of Sampling Distributions

-   **Center**: If samples are randomly selected, the sampling distribution will be centered around the **population parameter**.
-   **Shape**: For most of the statistics we consider, if the sample size is large enough the sampling distribution will be *symmetric and bell-shaped*.

**Caution**: These results only hold for random samples and will not apply if the sampling method is biased.

## Standard Error

-   To assess uncertainty of a statistic, we need to know its **sampling variability**, how much it varies from sample to sample.

-   **Standard error (SE)** is the *standard deviation* of a sample statistic.

-   The **standard error** measures how much the statistic varies from sample to sample.

-   The standard error of a statistic is just the **standard deviation of its sampling distribution.**

## What Happens to Sampling Variability as the Sample Size Gets Larger?

-   As the **sample size** *increases*, the variability of the sampling distribution *decreases*.

    -   As $n\uparrow$, the statistic gets closer to the corresponding parameter.

    -   As $n\uparrow$, $\text{SE}\downarrow$

## Exploring Sampling Distributions with [StatKey](https://www.lock5stat.com/StatKey/)

-   Introduction to StatKey: [Screencast 1](https://kzoo-my.sharepoint.com/:v:/g/personal/eric_nordmoe_kzoo_edu/EQfGXJC8eV1NjqXkCgxIBl0B3w0yMuENVi2rEZfaj78ckQ?e=nJRN6O)
-   Sampling distribution of the sample mean ($\bar X$): [Screencast 2](https://kzoo-my.sharepoint.com/:v:/g/personal/eric_nordmoe_kzoo_edu/EVvi_K5nODJBoNo_41Zm2tsBnzaaMvaoUsHVPh_6Yor7UA?e=T77BKH)
-   Sampling distribution of the sample proportion ($\hat p$)
    -   Built-in datasets: [Screencast 3](https://kzoo-my.sharepoint.com/:v:/g/personal/eric_nordmoe_kzoo_edu/ERbYIVME22VLsKUFxvxj-0ABidTsjVVB1MbE3X6yT5BswA?e=NXhsyf)
    -   Summary Statistics only: [Screencast 4](https://kzoo-my.sharepoint.com/:v:/g/personal/eric_nordmoe_kzoo_edu/EeSvinX7SvxDvYuPa_bpN9cBgbVIFaCIX_CnPfi7WuBNtw?e=TmfavQ)
