---
title: "4.2 Measuring Evidence With p-Values"
author: "E. Nordmoe"
date: "Math 260"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE)
```


```{r, echo=F,message=FALSE}
library(mosaic)
library(Lock5Data)
```





# Key Question

> How unusual is it to see a sample statistic as extreme as that observed, if $H_0$ were true?  

* If it is very unusual, we have statistically significant evidence against the null hypothesis  

$\Longrightarrow$ But *how* do we **measure** how unusual a sample statistic is, if $H_0$ were true?  

---
## Quantifying Evidence Against the Null Hypothesis  

> To see if a statistic provides evidence against $H_0$, we need to see what kind of sample statistics we would observe,  just by *random chance*, 
**if $H_0$ were true**  

???

Spinning pennies 20 times and count number of heads. Randomization distribution

---

# Randomization Distribution

> A **randomization distribution** is a collection of statistics from samples simulated assuming the null hypothesis is true  

--

* The randomization distribution shows what values of statistics would be observed, just by random chance, *if the null hypothesis were true*  

---

# Key Question

> How unusual is it to see a sample statistic as extreme as that observed, if $H_0$ is true?  

* A **randomization distribution** tells us what values of statistics we would see just by random chance, *if the null hypothesis were true*  

* Use the distribution to assess how extreme the observed statistic is! 

---

# p-value  

> The $p$-value is the chance of obtaining a sample statistic as extreme (or more extreme) than the observed sample statistic, *if the null hypothesis were true*  

--

* **Calculate** the $p$-value as the proportion of statistics in a randomization distribution that are as extreme (or more extreme) than the observed sample statistic  

---

# Calculating a p-value

1. What kinds of statistics would we get, just by random chance, if the null hypothesis were true?

  * Randomization distribution
--

2. What proportion of these statistics are as extreme as our original sample statistic?

  * $p$-value
    
---

# Alternative Hypothesis

* A one-sided alternative contains either $>$ or $<$   

* A two-sided alternative contains $\neq$  

* The $p$-value is the proportion **in the tail in the direction specified by $H_a$**  

* For a two-sided alternative, the $p$-value is **twice the proportion in the smallest tail**  


---

# Estimating a P-value from a Randomization Distribution  

* **One-tailed  alternative**:  Find  the  proportion  of  randomization
samples that **equal or exceed** the original statistic in the direction (tail)
indicated by the alternative hypothesis.  

--

* **Two-tailed  alternative**:  Find  the  proportion  of  randomization
samples in the smaller tail **at or beyond the original** statistic and then
**double the proportion** to account for the other tail.  

---

# Guidelines for Thinking About <br>p-values

When wording conclusions from a test of significance, you can consider:  

* $.05\le p \le .10 \implies$ *slight* evidence  

* $.01\le p \le .05 \implies$ *some* or *moderate* evidence  

* $p \le .01 \implies$ *strong* or *convincing* evidence

---

# Summary

* The randomization distribution shows what types of statistics would be observed, just by random chance, *if the null hypothesis were true*.

* A $p$-value is the chance of getting a statistic as extreme as that observed, *if $H_0$ were true*.

* A $p$-value can be calculated as the proportion of statistics in the randomization distribution *as extreme or more extreme than the observed sample statistic*.

* The smaller the *p*-value, the greater the evidence against $H_0$.