---
title: "4.4 Creating Randomization Distributions"
author: "E. Nordmoe"
date: "Math 260"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE)
```


```{r, echo=F,message=FALSE}
library(mosaic)
library(Lock5Data)
```


# Randomization Distributions  

$p$-values can be calculated by randomization distributions:  

* Simulate samples, assuming $H_0$ is true
* Calculate the statistic of interest for each sample  
* Find the $p$-value as the proportion of simulated statistics as extreme or more extreme than the observed statistic  

--

R can simulate randomization samples for many inference situations  

---

## Measuring Evidence Against the Null  

> To see if a statistic provides evidence against $H_0$, we need to see what kind of sample statistics we would observe,  just **by random chance**,  **if $H_0$ were true**  

---

# Randomization Distribution Center

A **randomization distribution** simulates samples assuming the null hypothesis is true, so  

* It is centered at the value of the parameter given in the null hypothesis.

--
* For differences, if the hypothesized value is 0, the distribution will be centered at 0.

---
    
# Randomization Distribution

The main criteria to consider when creating randomization samples for a statistical test are:

1. Be consistent with the null hypothesis $H_0$.  
    + If we don’t do this, we won’t be testing our null hypothesis.
--
2. Use the **observed data** in the original sample.  
    + We hope the original data will shed light on some aspects of the distribution that are not determined by null hypothesis.
--
3. Reflect the way the original data were collected.
    + Important but not our emphasis here.

**Note**: We'll consider several but not all situations and learn how to implement them manually and using R.

---

### Randomization Test Procedure Overview Using R  

1. State null and alternative hypotheses  

2. Use R to generate a randomization distribution 

3. Find and interpret the $p$-value  

4. State conclusion in context  

---

## Generating Randomization Samples

*  As long as the original data is used and the null hypothesis is true for the randomization samples, most methods usually give similar answers in terms of 
a $p$-value

*  R generates the randomization distribution for you, but you need to be able to interpret results in context with respect to the null and alternative hypotheses.

