---
title: "4.5 Making Connections"
author: "E. Nordmoe"
date: "Math 260"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE)
```


```{r, echo=F,message=FALSE}
library(mosaic)
library(Lock5Data)
```



# Bootstrap and Randomization Distributions

 Bootstrap Distributions | Randomization Distributions  
:----------------------------:|:------------------------------:
Our best guess at the distribution of sample statistics | Our best guess at the distribution of sample statistics, if $H_0$ were true   
Centered around the observed sample statistic | Centered around the null hypothesized value  
Simulate sampling from the population by resampling from the original sample |   Simulate samples assuming $H_0$ were true  


---
class:center, middle, inverse

# Key difference  

### Randomization distribution assumes $H_0$ true, <br> while a bootstrap distribution does *not*

---

# Intervals and Tests


* If a 95% CI **contains** the parameter in $H_0$, then a two-tailed test should **not reject** $H_0$ at the 5% significance level.

* If a 95% CI **misses** the parameter in $H_0$, then a two-tailed test should **reject** $H_0$ at the 5% significance level.

---

# Intervals and Tests

A confidence interval represents the **range of plausible values** for the population parameter

* If the null hypothesized value is **not within** the CI, it is **not a plausible** value and **should be rejected.**  

* If the null hypothesized value is **within** the CI, it **is a plausible** value and **should not be rejected.**  

---

# Intervals and Tests

* **Confidence intervals** give you a range of plausible values for the parameter.

* A **significance test** describes the plausibility of one specific value of the parameter (through the $p$-value).

---


# Summary

*  If a null hypothesized value lies **inside** a 95% CI, a two-tailed test using $\alpha=.05$ would **not reject** $H_0$.  

*  If a null hypothesized value lies **outside** a 95% CI, a two-tailed test using $\alpha=.05$ would **reject** $H_0$.  

* **Statistical significance** is not at all the same as **practical significance**. 


