---
title: "6.4 Inference About Differences in Means"
author: "E. Nordmoe"
date: "Math 260"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE)
```


```{r, echo=F,message=FALSE}
library(mosaic)
library(Lock5Data)
```


# Outline

* Standard Error   
* Sampling Distribution  
* Confidence Interval  
* Hypothesis Test  

---

# Context: Inference about the Difference in Means

* Comparison of means from two different (independent) populations.  
* Examples  
    + **Experiment**: Compare average blood pressure in treatment versus control group  
    + **Observational**: Compare average test scores for students from families below versus above the poverty line 
--
    
* BUT NOT  
    + Compare blood pressure before and after implementing an exercise regimen (This is a *matched pairs* comparison.)

---

# Standard Error for Difference in Means  

The standard error for $\bar x_1-\bar x_2$ is   

$$
SE=\sqrt{\frac{\sigma_1^2}{n_1}+\frac{\sigma_2^2}{n_2}}
$$  


* The larger the sample sizes, the smaller the SE  

---

# Central Limit Theorem for Difference in Means  

If $n_1$ and $n_2$ are sufficiently large,   

$$
\bar x_1-\bar x_2 \sim N\left(\mu_1-\mu_2, \sqrt{\frac{\sigma_1^2}{n_1}+\frac{\sigma_2^2}{n_2}}\right)
$$
--

Conditions:  

1. Independent random samples  
2. Population distributions are approximately normal OR sample sizes are large $(n_1\ge 30$ and $n_2\ge 30)$  

---

# Estimated Standard Error

Usually, we don’t know the population standard deviations $\sigma_1$ and $\sigma_2$, so we estimate them with the sample standard deviations to get the **standard error**  

$$
\text{SE}=\sqrt{\frac{s_1^2}{n_1}+\frac{s_2^2}{n_2}}
$$

---

# t-test for the Difference in Means  

Test $H_0: \mu_1=\mu_2$ using  

$$
t=\frac{\bar x_1-\bar x_2}{\sqrt{\frac{s_1^2}{n_1}+\frac{s_2^2}{n_2}}}
$$


---

# t-Distribution Confidence Interval for the Difference in Means  

A confidence interval for the difference in population means $\mu_1-\mu_2$ is given by:  

$$
(\bar x_1-\bar x_2) \pm t^* \sqrt{\frac{s_1^2}{n_1}+\frac{s_2^2}{n_2}}
$$

where $t^*$  is the critical value from the appropriate $t$-distribution.  

---

# Degrees of Freedom for t Approximation

Degrees of freedom for the $t$ distribution are:

*  Smaller of $n_1-1$ and $n_2-1$ when doing calculations by hand (conservative)

--

*  Given in the display below if done using R  or other technology (more accurate)

$$
{\text{df}} = \frac{{{{\left( {\frac{{s_1^2}}{{{n_1}}} + \frac{{s_2^2}}{{{n_2}}}} \right)}^2}}}{{\frac{1}{{{n_1} - 1}}{{\left( {\frac{{s_1^2}}{{{n_1}}}} \right)}^2} + \frac{1}{{{n_2} - 1}}\left( {\frac{{s_2^2}}{{{n_2}}}} \right)^2}}
$$

---

# Conditions for t-Inference for the Difference in Means

Conditions: 

1. Independent random samples

2. Population distributions are approximately normal OR sample sizes are large $(n_1\ge 30$ and $n_2\ge 30)$


