---
title: "6.5 Paired Difference in Means"
author: "E. Nordmoe"
date: "Math 260"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE)
```


```{r, echo=F,message=FALSE}
library(mosaic)
library(Lock5Data)
```


# Outline

* Paired data     
* Analyzing paired data  
* Inference for paired data
* Conditions

---

# Context: Inference about the Difference in Means for Paired Data

**Paired data** includes the following:    

* Two measurements on each unit    
    + Same unit measured under different experimental conditions  
* Measurements on a pair of units that go together
    + Twin studies  
    + Partner studies

---

# Analyzing Paired Data: <br> One-Sample t


* For a matched pairs experiment, we look at the difference between responses for each unit (pair), rather than just the average difference between treatment groups.  

* Get a new variable of the differences, and do inference for the difference as you would for a single mean.  

* Rather than doing inference for a difference in means, do inference for the mean difference.  

---

# Inference for Paired Data

To analyze the differences, we use the same formulas we already learned for a single mean:

|            Standard Error           |            Confidence Interval            |                  Hypothesis Test                 |
|:-----------------------------------:|:-----------------------------------------:|:------------------------------------------------:|
| $\text{SE}= \frac{s_d}{\sqrt{n_d}}$ | $\bar x_d \pm t^* \frac{s_d}{\sqrt{n_d}}$ | $t=\displaystyle\frac{\bar x_d}{s_d/\sqrt{n_d}}$ |
  
  
where  

* $\bar x_d$ is the average difference  
* $s_d$ is the standard deviation of the differences and  
* $n_d$ is the number of differences (pairs)

---

# Conditions for t-Inference for Matched Pairs

Conditions:  

1. Simple random sample of differences
2. Population of differences is approximately normal OR sample size of differences is  large $(n_d\ge 30)$


