---
title: "6.3 Inference About Differences in Proportions"
author: "E. Nordmoe"
date: "Math 260"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE)
```


```{r, echo=F,message=FALSE}
library(mosaic)
library(Lock5Data)
```


# Outline
* Standard Error   
* Sampling Distribution  
* Confidence Interval  
* Hypothesis Test  

---

# Context: Inference about the Difference in Proportions  

Comparison of proportions from two different (independent) populations.  

Examples  

  * Compare success proportion in Group A versus Group B   
  * Compare survival for group getting Treatment A versus survival for group getting Treatment B   

--

BUT NOT 

  * Compare proportion of voters who prefer A to proportion who prefer B  
    + This is a comparison of proportions within the same group.  

---

# Standard Error for Difference in Proportions  

The standard error for $\hat p_1-\hat p_2$ is
$$
\text{SE}=\sqrt{\frac{p_1(1-p_1)}{n_1}+\frac{p_2(1-p_2)}{n_2}}
$$

---

# Central Limit Theorem for Difference in Proportions  

If $n$ is sufficiently large:   

$$
\hat p_1-\hat p_2 \sim N\left(p_1-p_2, \sqrt{\frac{p_1(1-p_1)}{n_1}+\frac{p_2(1-p_2)}{n_2}}\right)
$$

Conditions:  

  * Independent random samples  
  * Sufficiently large samples  
    + $n_1 p_1\ge 10$ and $n_1(1-p_1)\ge 10$  
    + $n_2 p_2\ge 10$ and $n_2(1-p_2)\ge 10$  

---

# Confidence Interval for a Difference in Proportions  


A confidence interval for $p_1-p_2$ can be computed as  

$$
(\hat p_1-\hat p_2) \pm z^* \sqrt{\frac{\hat p_1(1-\hat p_1)}{n_1}+\frac{\hat p_2(1-\hat p_2)}{n_2}}
$$

Conditions:  

  * Independent random samples  
  * Sufficiently large samples  
    + At least 10 successes and failures in each group  

---

# Test for Difference in Proportions  

To test $H_0:p_1=p_2$, based on samples of size $n_1$ and $n_2$ from two groups, calculate  
$$
z=\frac{\hat p_1-\hat p_2}{\sqrt{\frac{\hat p(1-\hat p)}{n_1}+\frac{\hat p(1-\hat p)}{n_2}}}
$$
where $\hat p$ is the *pooled* proportion obtained by combining the two samples.  That is,  

$$
\hat p=\frac{\text{Total Successes}}{\text{Total Sample Size}}
$$

---

# Test for Difference in Proportions  (continued)  

Conditions:  

  * Independent random samples  
  * Sufficiently large samples  
    + At least 10 successes and failures in each group  

---

# Test for Difference in Proportions: Simplified Formula

The $z$-test statistic can be simplified by factoring:  

$$
z=\frac{\hat p_1-\hat p_2}{\sqrt{\hat p(1-\hat p)\left(\frac{1}{n_1}+\frac{1}{n_2}\right)}}
$$
where $\hat p$ is again the *pooled* proportion obtained by combining the two samples.  


