---
title: "7.1 Testing Goodness-of-Fit for a Single Categorical Variable"
author: "E. Nordmoe"
date: "Math 260"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE)
```


```{r, echo=F,message=FALSE}
library(mosaic)
library(Lock5Data)
```


# Outline

* Test for a single categorical variable  
    + Hypotheses  
    + Expected counts  
    + Chi-square statistic  
    + Chi-square distribution  

---

# Multiple Categories  

* We know how to test a proportion for a single category of a categorical variable.  
    + Example: Test if the proportion of times a die lands 1 is different from 1/6.  
    
* What if we want to test proportions for several categories at once?   
    + Are the 6 outcomes of a die equally likely?  
    
---

# Chi-square Statistic

The **chi-square statistic**  compares the *observed* to *expected* counts for derived from a null hypothesis:  
$$
\chi^2=\sum \frac{(\text{Observed}-\text{Expected})^2}{\text{Expected}}
$$

where the sum is over all cells of the table.  

---

# P-value for a Chi-Square Statistic  

Two options for a $p$-value:  

1. Randomization test (Using StatKey)
    + See More Advanced Randomization Tests row
--

2. Chi-square distribution  
    + degrees of freedom = number of categories-1
    

---

# Chi-square Distribution

* If each of the expected counts are at least 5, AND  

* If the null hypothesis is true, then the $\chi^2$  statistic follows a $\chi^2$ distribution, with degrees of freedom equal to  

$$
\text{df = number of categories – 1}
$$

--
<br><br><br>
**Note:** The $p$-value is always one-sided (right-tailed) since only *large* values of $\chi^2$ are evidence *against* the hypothesized proportions.  

---

# Goodness-of-Fit

* A $\chi^2$ test for goodness-of-fit tests whether the distribution of a categorical variable is the same as some *null hypothesized 
distribution*.  

* The null hypothesized proportions for each category do not have to be the same.  

---

# Chi-Square Test for Goodness of Fit  

1. State null hypothesized proportions for each category, $p_i$.
    + Alternative is that at least one of the proportions is different than specified in the null.  
2. Calculate the expected counts for each cell as $np_i$.
    + Make sure they are all greater than 5 to proceed.

3. Calculate the $\chi^2$ statistic:
$$
\chi^2=\sum \frac{(\text{Observed}-\text{Expected})^2}{\text{Expected}}
$$

---

# Chi-Square Test for Goodness of Fit (cont'd)

<span>4.</span> Compute the $p$-value 
as the area in the tail above the $\chi^2$ statistic, for a $\chi^2$ distribution with  df = (# of categories – 1).

<span>5.</span> Interpret the $p$-value in context.

---

# Summary


* Use a chi-square goodness-of-fit test to assess hypotheses about the distribution of categories for a single categorical variable.   

* Data are observed counts in each category (cell) that we compare to expected counts (under $H_0$) using  

$$
\chi^2=\sum \frac{(\text{Observed}-\text{Expected})^2}{\text{Expected}}.
$$

* Find a $p$-value using the right (upper) tail of a chi-square distribution  with df = (# of categories – 1).  
