---
title: "7.2 Testing for an Association Between Two Categorical Variables"
author: "E. Nordmoe"
date: "Math 260"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE)
```


```{r, echo=F,message=FALSE}
library(mosaic)
library(Lock5Data)
```


# Outline

* Test for a relationship between two categorical variables  
    + Hypotheses  
    + Expected counts  
    + Chi-square test for association (independence)   

---

#Recap: Chi-square Goodness-of-Fit Test  

* Null hypothesis specifies proportions $p_i$ for each of $k$ categories of a *single* categorical variable.   
* Data= *observed counts* in each category (from a sample of size $n$).  
* Compare *expected counts*= $np_i$ for each cell to observed counts using  

$$
\chi^2=\sum \frac{(\text{Observed}-\text{Expected})^2}{\text{Expected}}.
$$

* Find a $p$-value using the right (upper) tail of a chi-square distribution with df = (# of categories – 1).  

---

# Two Categorical Variables

* The statistics behind a Chi-square test easily extends to *two* categorical variables.

* A **chi-square test for association** (often called a **chi-square test for independence**) test for an *association* between two categorical variables.

---

# Chi-square Distribution for a Two-Way Table

Use the same chi-square statistic:

$$
\chi^2=\sum \frac{(\text{Observed}-\text{Expected})^2}{\text{Expected}}.
$$

* For testing an $r\times c$ table, we can find a $p$-value using the right (upper) tail of a $\chi^2$-distribution
  + With $(r-1)(c-1)$ degrees of freedom *if*
  
  + *Expected cell counts* are at least  5 in all cells.
    
* Notation: $r=$ number of rows and $c=$ number of columns in table.

* If expected cell counts are not met, a randomization test in R could be used.

---

# Interpreting Results of the Chi-Square Test

* The chi-square test only indicates **if** the variables are associated.

* The **contribution** in each cell helps understand the nature of the relationship.

---

# Chi-Square Test for Association

1. Test $H_0:$ the two variables are not associated vs $H_a:$ the two variables are associated.

2. Calculate the expected counts for each cell. Check that all expected are at least 5 to proceed.
$$
\text{Expected cell count}=\frac{\text{Row Total }\times \text{ Column Total}}{\text{Overall Total}}
$$

3. Calculate the $X^2$ statistic:
$$
X^2=\sum \frac{(\text{Observed}-\text{Expected})^2}{\text{Expected}}.
$$

---

# Chi-Square Test for Association (continued)  

<span>4.</span> Compute the p-value as the area in the tail above the $X^2$ statistic, using a $\chi^2$-distribution with $\text{df} = (r – 1) \times (c – 1)$.

<span>5.</span> Interpret the $p$-value in context.

---

# Chi-square and Z-test for 2 by 2 Table  

* When testing 2 categorical variables with 2 categories each $(2\times 2$ table), the $\chi^2$ statistic is exactly the square of the $z$-statistic:  

$$
\chi^2=z^2
$$

* The $X^2$ distribution with 1 df is exactly the square of the standard normal distribution.  
* Doing a two-sided $z$-test for difference in proportions or doing a chi-square test for a $2\times 2$ table will produce exactly the same $p$-value.  

---

# Summary Chi-square Tests  

* The **chi-square test for goodness-of-fit** tests if *one* categorical variable differs from a hypothesized distribution.  
* The **chi-square test for association** tests if *two* categorical variables are associated.  
* For both, you calculate expected counts for each cell and compute   

$$
\chi^2=\sum \frac{(\text{Observed}-\text{Expected})^2}{\text{Expected}}.
$$

* Find the $p$-value as the area above this statistic with a $\chi^2$ distribution using degrees of freedom   
    + $k-1$ for test for goodness-of-fit  
    + $(r-1)\times(c-1)$ for test for association.


