---
title: "Investigative Group Worksheet: Bootstrap t Intervals"
author: "Mathematical Statistics with Resampling"
date: "2024-05-17"
format: html
---

# Investigative Group Worksheet: Bootstrap t Intervals

## Objectives
- Understand and implement Bootstrap t intervals for a single mean.
- Understand and implement Bootstrap t intervals for the difference in means.
- Interpret the results and discuss their implications.

## Instructions
Work in groups to complete the following tasks. Use R and the provided code snippets to guide your analysis. Discuss your findings with your group and prepare to share your results with the class.

### Part 1: Bootstrap t Interval for a Single Mean

#### Task 1: Load the Data
Load the `mtcars` dataset and focus on the `mpg` (miles per gallon) variable.

```{r}
data(mtcars)
mpg <- mtcars$mpg
```

#### Task 2: Calculate the Bootstrap t Interval
Write a function to calculate the Bootstrap t interval for the mean of `mpg`.

```{r}
set.seed(123)
bootstrap_t_interval <- function(data, B = 1000, alpha = 0.05) {
  n <- length(data)
  original_mean <- mean(data)
  bootstrap_means <- numeric(B)
  bootstrap_se <- numeric(B)
  
  for (i in 1:B) {
    sample_data <- sample(data, n, replace = TRUE)
    bootstrap_means[i] <- mean(sample_data)
    bootstrap_se[i] <- sd(sample_data) / sqrt(n)
  }
  
  t_star <- qt(1 - alpha / 2, df = n - 1)
  lower <- original_mean - t_star * sd(bootstrap_means)
  upper <- original_mean + t_star * sd(bootstrap_means)
  
  return(c(lower, upper))
}

interval <- bootstrap_t_interval(mpg)
interval
```

#### Task 3: Interpret the Results
1. What are the lower and upper bounds of the 95% Bootstrap t interval for the mean of `mpg`?
2. How do these results compare to the traditional t-interval obtained using the `t.test` function?

```{r}
t.test(mpg)$conf.int
```

### Part 2: Bootstrap t Interval for the Difference in Means

#### Task 4: Load the Data
Split the `mtcars` dataset into two groups based on the number of cylinders (`cyl`): 4-cylinder and 6-cylinder cars. Focus on the `mpg` variable.

```{r}
mpg_4cyl <- mtcars$mpg[mtcars$cyl == 4]
mpg_6cyl <- mtcars$mpg[mtcars$cyl == 6]
```

#### Task 5: Calculate the Bootstrap t Interval for the Difference in Means
Write a function to calculate the Bootstrap t interval for the difference in means between 4-cylinder and 6-cylinder cars.

```{r}
bootstrap_diff_t_interval <- function(data1, data2, B = 1000, alpha = 0.05) {
  n1 <- length(data1)
  n2 <- length(data2)
  original_diff <- mean(data1) - mean(data2)
  bootstrap_diffs <- numeric(B)
  bootstrap_se <- numeric(B)
  
  for (i in 1:B) {
    sample_data1 <- sample(data1, n1, replace = TRUE)
    sample_data2 <- sample(data2, n2, replace = TRUE)
    bootstrap_diffs[i] <- mean(sample_data1) - mean(sample_data2)
    bootstrap_se[i] <- sqrt(sd(sample_data1)^2 / n1 + sd(sample_data2)^2 / n2)
  }
  
  t_star <- qt(1 - alpha / 2, df = min(n1, n2) - 1)
  lower <- original_diff - t_star * sd(bootstrap_diffs)
  upper <- original_diff + t_star * sd(bootstrap_diffs)
  
  return(c(lower, upper))
}

diff_interval <- bootstrap_diff_t_interval(mpg_4cyl, mpg_6cyl)
diff_interval
```

#### Task 6: Interpret the Results
1. What are the lower and upper bounds of the 95% Bootstrap t interval for the difference in means between 4-cylinder and 6-cylinder cars?
2. How do these results compare to the traditional t-interval obtained using the `t.test` function?

```{r}
t.test(mpg_4cyl, mpg_6cyl)$conf.int
```

### Discussion Questions
1. Discuss the advantages and limitations of using Bootstrap t intervals compared to traditional methods.
2. How might the sample size affect the width of the Bootstrap t intervals?
3. What assumptions are we making when using Bootstrap t intervals? Are these assumptions reasonable in the context of this dataset?

## Submission
Submit your group's findings, including the R code and interpretations, to the class Moodle page.

---

Happy Bootstrapping!
